---
title: "WTS - Quantification Result Comparison"
date: "`r format(Sys.time(), '%a %Y-%b-%d')`"
output:
  html_document:
    keep_md: yes
    css: style.css
    theme: readable
    toc: true
    toc_float: true
    code_folding: hide
  rmdformats::material:
    highlight: kate
params:
  run_1: '~/UMCCR/data/projects/WTS_comparison/dragen_4.2_4.4/SBJ01563/dragen_4.2'
  run_1_name: 'dragen_4.2'
  run_2: '~/UMCCR/data/projects/WTS_comparison/dragen_4.2_4.4/SBJ01563/dragen_4.4'
  run_2_name: 'dragen_4.4'
---

## Introduction

<div class = "blue">
This document compares WTS workflow abundance outputs, reporting:

- File names in the input (params) directories
- The number and comparison of calls
- Key differences between both outputs in a tabular format
</div>


## List files

This section lists the file names for outputs in the run directories being compared in this report.  

```{r read_data, message=FALSE, warning=FALSE}
###Read input. 
###The script expects two input directories, each containing 'final' results as produced by bcbio. The results in these two input directories will be compared.
run_1 <-file.path(params$run_1)
#list files in the first run directory
list.files(run_1)

run_2 <- file.path(params$run_2)
#list files in the second run directory
list.files(run_2)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# Load required libraries
library(tools)
library(here)
library(DT)
library(dplyr)
library(ggplot2)
library(plotly)
library(gridExtra)
library(tximport)
library(ggpubr)
library(tidyr)
```

```{r funcs, include=FALSE}
# Function to find and read quant.genes.sf file
read_quant_file <- function(dir) {
  file <- list.files(dir, pattern = "quant.genes.sf$", recursive = TRUE, full.names = TRUE)
  if (length(file) == 0) stop("No quant.genes.sf file found in ", dir)
  data <- read.table(file[1], sep="\t", as.is=TRUE, header=TRUE)
  data$Name <- sub("\\..*", "", data$Name)
  return(data)
}
```

## Data Preparation

```{r input-counts, message=FALSE, warning=FALSE}
# Read in the data
counts_1 <- read_quant_file(params$run_1)
counts_2 <- read_quant_file(params$run_2)

# Create variable names from file names
var_name_1 <- tools::file_path_sans_ext(basename(list.files(params$run_1, pattern = "quant.genes.sf$", recursive = TRUE)[1]))
var_name_2 <- tools::file_path_sans_ext(basename(list.files(params$run_2, pattern = "quant.genes.sf$", recursive = TRUE)[1]))

# Assign to dynamically named variables
assign(var_name_1, counts_1)
assign(var_name_2, counts_2)

# Remove version numbers from gene names
counts_1$Name <- sub("\\..*", "", counts_1$Name)
counts_2$Name <- sub("\\..*", "", counts_2$Name)

# Join the datasets
#combine.counts <- inner_join(counts_1, counts_2 , by = "Name", suffix = c(run_1_name, run_2_name)) %>%
#  dplyr::mutate(percentageDiff = ifelse(TPM_run1 > TPM_run2, (TPM_run1 - TPM_run2)/((TPM_run1 + TPM_run2)/2) *100, (TPM_run2 - TPM_run1)/((TPM_run1 + TPM_run2)/2) *100)) %>%
#  dplyr::mutate_if(is.numeric, round,2)

# Read params
run_1_name <- params$run_1_name
run_2_name <- params$run_2_name

combine.counts <- inner_join(counts_1, counts_2, by = "Name", suffix = c(paste0("_", run_1_name), paste0("_", run_2_name))) %>%
  dplyr::mutate(percentageDiff = ifelse(
    .data[[paste0("TPM_", run_1_name)]] > .data[[paste0("TPM_", run_2_name)]],
    (.data[[paste0("TPM_", run_1_name)]] - .data[[paste0("TPM_", run_2_name)]]) / 
      ((.data[[paste0("TPM_", run_1_name)]] + .data[[paste0("TPM_", run_2_name)]]) / 2) * 100,
    (.data[[paste0("TPM_", run_2_name)]] - .data[[paste0("TPM_", run_1_name)]]) / 
      ((.data[[paste0("TPM_", run_1_name)]] + .data[[paste0("TPM_", run_2_name)]]) / 2) * 100
  )) %>%
  dplyr::mutate_if(is.numeric, round, 2)

# Load gene names
tx_gene_id_105_rds <- system.file("extdata/ensembl/tx_gene_id_105.rds", package = "RNAsum.data")
tx_gene_id_105 <- readr::read_rds(tx_gene_id_105_rds) %>%
  dplyr::select(GENEID, GENENAME) %>%
  dplyr::rename(NAME = GENEID)
tx_gene_id_105_unique <- tx_gene_id_105 %>%
  distinct(NAME, .keep_all = TRUE)

# Prepare the main dataset
est.counts <- combine.counts[, c('Name', paste0('TPM_', run_1_name), paste0('TPM_', run_2_name), 'percentageDiff')] %>%
  left_join(tx_gene_id_105_unique, by = c("Name" = "NAME")) %>%
  dplyr::rename(Gene_name = GENENAME)

# Filter data - keep difference above third quantile and remove lowly expressed genes
est.counts.filtered <- est.counts %>%
  filter(!is.na(percentageDiff)) %>%  # Remove NA values before calculating quantile
  filter(percentageDiff > quantile(percentageDiff, 0.75, na.rm = TRUE)) %>%
  filter(paste0('TPM_', run_1_name) > 1 | paste0('TPM_', run_2_name) > 1)
```

## Analysis Results

### Total Gene Counts

#### Summary Table

```{r summary, message=FALSE, warning=FALSE}
lines.count <- data.frame(c(run_1_name, run_2_name), c(nrow(counts_1), nrow(counts_2)))
colnames(lines.count) <- c("compare", "lines_count")
datatable(lines.count, rownames = TRUE, options = list(sDom = '<"top">lrt<"bottom">ip'))
```

### Gene Expression Correlation

#### Correlation Plot

```{r cor, message=FALSE, warning=FALSE}
cor.gene <- cor.test(log2(est.counts[[paste0("TPM_", run_1_name)]] + 1),
                     log2(est.counts[[paste0("TPM_", run_2_name)]] + 1),
                     method = 'spearman')

ggplot(est.counts,
       aes(x = log2(.data[[paste0('TPM_', run_1_name)]] + 1), 
           y = log2(.data[[paste0('TPM_', run_2_name)]] + 1))) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "red") +
  stat_cor(method = "spearman", label.x = 0, label.y = 20) +
  labs(title = "Per gene counts correlation",
       x = paste0("log2(TPM_", run_1_name, " + 1)"),
       y = paste0("log2(TPM_", run_2_name, " + 1)")) +
  theme_minimal() +
  coord_fixed(ratio = 1)
```

### TPM Distribution Analysis

#### TPM Summary Statistics

The large difference between the median and mean below indicates a highly skewed distribution of gene expression. A few genes are very highly expressed, while most genes have low or no expression.

Third quarile value was used as a TPM threshold to filter out lowly expressed genes as 75% of genes have TPM ≤ to this value.

Also, the data is log-transformed for analyses and visualizations to better handle the wide dynamic range.

```{r sum-stats, message=FALSE, warning=FALSE}
# Calculate summaries
summary_run1 <- summary(est.counts[[paste0("TPM_", run_1_name)]])
summary_run2 <- summary(est.counts[[paste0("TPM_", run_2_name)]])

# Create summary statistics dataframe
summary_stats <- data.frame(
  Statistic = names(summary_run1),
  Run1 = as.numeric(summary_run1),
  Run2 = as.numeric(summary_run2)
) %>%
  mutate(across(where(is.numeric), ~round(., 2)))

# Rename columns dynamically
colnames(summary_stats)[2:3] <- c(paste0("TPM_", run_1_name), paste0("TPM_", run_2_name))

# Create datatable
datatable(summary_stats,
          options = list(
            pageLength = 6,
            dom = 't',
            ordering = FALSE
          ),
          rownames = FALSE,
          caption = paste("Summary Statistics of TPM Values for", run_1_name, "and", run_2_name)) %>%
  formatStyle(columns = c(paste0("TPM_", run_1_name), paste0("TPM_", run_2_name)),
              backgroundColor = styleInterval(c(1, 10, 100),
                                              c('white', '#f7f7f7', '#e7e7e7', '#d7d7d7')))
```

#### TPM Distribution Plots

```{r tpm, message=FALSE, warning=FALSE}
# Create visualisation
threshold <- 1

# Prepare the data
plot_data <- est.counts.filtered %>%
  select(!!sym(paste0("TPM_", run_1_name)), !!sym(paste0("TPM_", run_2_name))) %>%
  pivot_longer(cols = everything(), names_to = "TPM_type", values_to = "TPM_value") %>%
  mutate(TPM_type = factor(TPM_type, 
                           levels = c(paste0("TPM_", run_1_name), paste0("TPM_", run_2_name)),
                           labels = c(run_1_name, run_2_name)))

# Create the plot
ggplot(plot_data, aes(x = log2(TPM_value))) +
  geom_histogram(bins = 80, fill = "skyblue", color = "black") +
  scale_x_continuous(limits = c(0, 12)) +
  facet_wrap(~ TPM_type, scales = "free_y", ncol = 2) +
  labs(title = paste("Distribution of log2(TPM) for Filtered Data (TPM >", threshold, ")"),
       x = "log2(TPM)",
       y = "Count") +
  theme_minimal() +
  theme(strip.background = element_rect(fill = "lightgray"),
        strip.text = element_text(face = "bold"))
```

#### Display filtered data (TPM > 1) and TPM scatter plot

```{r tpm-scatter, message=FALSE, warning=FALSE}
# Display filtered data
datatable(est.counts.filtered, 
          rownames = FALSE, 
          filter = "top", 
          options = list(pageLength = 10, scrollX = TRUE), 
          caption = "Filtered gene counts (values above third quartile)")

p <- plot_ly(est.counts.filtered, 
             x = ~log2(.data[[paste0("TPM_", run_1_name)]] + 1), 
             y = ~log2(.data[[paste0("TPM_", run_2_name)]] + 1), 
             type = 'scatter', 
             mode = 'markers',
             color = ~percentageDiff,
             colors = colorRamp(c("blue", "white", "red")),
             marker = list(
               colorbar = list(title = "% Difference"),
               colorscale = 'RdBu',
               cmin = 50,
               cmid = 100,
               cmax = 200
             ),
             text = ~paste('Gene Name:', Gene_name, 
                           '<br>', run_1_name, ' TPM:', round(.data[[paste0("TPM_", run_1_name)]], 2), 
                           '<br>', run_2_name, ' TPM:', round(.data[[paste0("TPM_", run_2_name)]], 2),
                           '<br>Percentage Diff:', round(percentageDiff, 2)),
             hoverinfo = 'text') %>% 
  layout(title = paste("Scatter plot for filtered counts<br>Percentage difference above third quartile"),
         xaxis = list(title = paste0("log2(TPM_", run_1_name, " + 1)")),
         yaxis = list(title = paste0("log2(TPM_", run_2_name, " + 1)")))

# Display plot
p
```
  
