---
title: "WTS - Quantification Result Comparison"
date: "`r format(Sys.time(), '%a %Y-%b-%d')`"
output:
  html_document:
    keep_md: yes
    css: style.css
    theme: readable
    toc: true
    toc_float: true
    code_folding: hide
  rmdformats::material:
    highlight: kate
params:
  run_1: '~/UMCCR/research/projects/nebVSillumina/SBJ03118/neb'
  run_2: '~/UMCCR/research/projects/nebVSillumina/SBJ03118/illumina'
---

## Introduction

<div class = "blue">
This document compares WTS workflow abundance outputs, reporting:

- File names in the input (params) directories
- The number and comparison of calls
- Key differences between both outputs in a tabular format
</div>


## List files

This section lists the file names for outputs in the run directories being compared in this report.  

```{r read_data, message=FALSE, warning=FALSE}
###Read input. 
###The script expects two input directories, each containing 'final' results as produced by bcbio. The results in these two input directories will be compared.
run_1 <-file.path(params$run_1)
#list files in the first run directory
list.files(run_1)

run_2 <- file.path(params$run_2)
#list files in the second run directory
list.files(run_2)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# Load required libraries
library(tools)
library(here)
library(DT)
library(dplyr)
library(ggplot2)
library(plotly)
library(gridExtra)
library(tximport)
library(ggpubr)
library(tidyr)
```

```{r funcs, include=FALSE}
# Function to find and read quant.genes.sf file
read_quant_file <- function(dir) {
  file <- list.files(dir, pattern = "quant.genes.sf$", recursive = TRUE, full.names = TRUE)
  if (length(file) == 0) stop("No quant.genes.sf file found in ", dir)
  data <- read.table(file[1], sep="\t", as.is=TRUE, header=TRUE)
  data$Name <- sub("\\..*", "", data$Name)
  return(data)
}
```

## Data Preparation

```{r input-counts, message=FALSE, warning=FALSE}
# Read in the data
counts_1 <- read_quant_file(params$run_1)
counts_2 <- read_quant_file(params$run_2)

# Create variable names from file names
var_name_1 <- tools::file_path_sans_ext(basename(list.files(params$run_1, pattern = "quant.genes.sf$", recursive = TRUE)[1]))
var_name_2 <- tools::file_path_sans_ext(basename(list.files(params$run_2, pattern = "quant.genes.sf$", recursive = TRUE)[1]))

# Assign to dynamically named variables
assign(var_name_1, counts_1)
assign(var_name_2, counts_2)

# Remove version numbers from gene names
counts_1$Name <- sub("\\..*", "", counts_1$Name)
counts_2$Name <- sub("\\..*", "", counts_2$Name)

# Join the datasets
neb.ilumina.counts <- inner_join(counts_1, counts_2 , by = "Name", suffix = c("_neb", "_illumina")) %>%
  dplyr::mutate(percentageDiff = ifelse(TPM_neb > TPM_illumina, (TPM_neb - TPM_illumina)/((TPM_neb + TPM_illumina)/2) *100, (TPM_illumina - TPM_neb)/((TPM_neb + TPM_illumina)/2) *100)) %>%
  dplyr::mutate_if(is.numeric, round,2)

# Load gene names
tx_gene_id_105_rds <- system.file("extdata/ensembl/tx_gene_id_105.rds", package = "RNAsum.data")
tx_gene_id_105 <- readr::read_rds(tx_gene_id_105_rds) %>%
  dplyr::select(GENEID, GENENAME) %>%
  dplyr::rename(NAME = GENEID)
tx_gene_id_105_unique <- tx_gene_id_105 %>%
  distinct(NAME, .keep_all = TRUE)

# Prepare the main dataset
est.counts <- neb.ilumina.counts[, c('Name', 'TPM_neb', 'TPM_illumina', 'percentageDiff')] %>%
  left_join(tx_gene_id_105_unique, by = c("Name" = "NAME")) %>%
  dplyr::rename(Gene_name = GENENAME)

# Filter data - keep difference above third quantile and remove lowly expressed genes
est.counts.filtered <- est.counts %>%
  filter(!is.na(percentageDiff)) %>%  # Remove NA values before calculating quantile
  filter(percentageDiff > quantile(percentageDiff, 0.75, na.rm = TRUE)) %>%
  filter(TPM_neb > 1 | TPM_illumina > 1)
```

## Analysis Results

### Total Gene Counts

#### Summary Table

```{r summary, message=FALSE, warning=FALSE}
lines.count <- data.frame(c("NEB", "ILLUMINA"), c(nrow(counts_1), nrow(counts_2)))
colnames(lines.count) <- c("assay", "lines_count")
datatable(lines.count, rownames = TRUE, options = list(sDom = '<"top">lrt<"bottom">ip'))
```

### Gene Expression Correlation

#### Correlation Plot

```{r cor, message=FALSE, warning=FALSE}
cor.gene <- cor.test(log2(est.counts$TPM_neb + 1),
                     log2(est.counts$TPM_illumina + 1),
                     method = 'spearman')

ggplot(est.counts,
       aes(x = log2(TPM_neb + 1), y = log2(TPM_illumina + 1))) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "red") +
  stat_cor(method = "spearman", label.x = 0, label.y = 20) +
  labs(title = "Per gene counts correlation between NEB and Illumina",
       x = "log2(TPM_neb + 1)",
       y = "log2(TPM_illumina + 1)") +
  theme_minimal() +
  coord_fixed(ratio = 1)
```

### TPM Distribution Analysis

#### TPM Summary Statistics

The large difference between the median (0.00) and mean (16.23) below indicates a highly skewed distribution of gene expression. A few genes are very highly expressed, while most genes have low or no expression.

Given that 75% of genes have TPM ≤ 1.13, a TPM threshold (1 TPM) was used to filter out lowly expressed genes.

Also, the data is log-transformed for analyses and visualizations to better handle the wide dynamic range.

```{r sum-stats, message=FALSE, warning=FALSE}
summary_neb <- summary(est.counts$TPM_neb)
summary_illumina <- summary(est.counts$TPM_illumina)

summary_stats <- data.frame(
  Statistic = names(summary_neb),
  TPM_neb = as.numeric(summary_neb),
  TPM_illumina = as.numeric(summary_illumina)
) %>%
  mutate(across(where(is.numeric), ~round(., 2)))

datatable(summary_stats,
          options = list(
            pageLength = 6,
            dom = 't',
            ordering = FALSE
          ),
          rownames = FALSE,
          caption = "Summary Statistics of TPM Values") %>%
  formatStyle(columns = c('TPM_neb', 'TPM_illumina'),
              backgroundColor = styleInterval(c(1, 10, 100),
                                              c('white', '#f7f7f7', '#e7e7e7', '#d7d7d7')))
```

#### TPM Distribution Plots

```{r tpm, message=FALSE, warning=FALSE}
# Create visualisation
threshold <- 1

# Prepare the data
plot_data <- est.counts.filtered %>%
  select(TPM_neb, TPM_illumina) %>%
  pivot_longer(cols = c(TPM_neb, TPM_illumina), names_to = "TPM_type", values_to = "TPM_value")

# Create the plot
ggplot(plot_data, aes(x = log2(TPM_value))) +
  geom_histogram(bins = 80, fill = "skyblue", color = "black") +
  scale_x_continuous(limits = c(0, 12)) +
  facet_wrap(~ TPM_type, scales = "free_y", ncol = 2) +
  labs(title = "Distribution of log2(TPM) for Filtered Data (TPM > 1)",
       x = "log2(TPM)",
       y = "Count") +
  theme_minimal() +
  theme(strip.background = element_rect(fill = "lightgray"),
        strip.text = element_text(face = "bold"))
```

#### Display filtered data (TPM >1) and TPM scatter plot

```{r tpm-scatter, message=FALSE, warning=FALSE}
# Display filtered data
datatable(est.counts.filtered, 
          rownames = FALSE, 
          filter = "top", 
          options = list(pageLength = 10, scrollX = TRUE), 
          caption = "Filtered gene counts (values above third quartile)")

p <- plot_ly(est.counts.filtered, 
             x = ~log2(TPM_neb + 1), 
             y = ~log2(TPM_illumina + 1), 
             type = 'scatter', 
             mode = 'markers',
             color = ~percentageDiff,
             colors = colorRamp(c("blue", "white", "red")),
             marker = list(
               colorbar = list(title = "% Difference"),
               colorscale = 'RdBu',
               cmin = 50,
               cmid = 100,
               cmax = 200
             ),
             text = ~paste('Gene Name:', Gene_name, 
                           '<br>Neb TPM:', round(TPM_neb, 2), 
                           '<br>Illumina TPM:', round(TPM_illumina, 2),
                           '<br>Percentage Diff:', round(percentageDiff, 2)),
             hoverinfo = 'text') %>% 
  layout(title = "Scatter plot for filtered counts<br>Percentage difference above third quartile",
         xaxis = list(title = "log2(TPM_neb + 1)"),
         yaxis = list(title = "log2(TPM_illumina + 1)"))

# Display plot
p
```

